---
title: "Data download and preparation for study ST000291"
author: "Alex Sanchez"
format: html
---

```{r echo=FALSE, message=FALSE}
if (!require(metabolomicsWorkbenchR)) 
  BiocManager::install("metabolomicsWorkbenchR")
if (!require(fobitools)) 
  BiocManager::install("fobitools")
if (!require(SummarizedExperiment)) 
  BiocManager::install("SummarizedExperiment")

if (!require(dplyr)) install.packages("dplyr", dep=TRUE)
if (!require(rvest)) install.packages("rvest", dep=TRUE)
if (!require(xlsx)) install.packages("xlsx", dep=TRUE)

```
# Data download

This study is composed of two complementary MS analyses, 
the positive mode (AN000464) and the negative mode (AN000465). 
Both are downloaded


```{r}
library(metabolomicsWorkbenchR)
data_negative_mode <- do_query(
  context = "study",
  input_item = "analysis_id",
  input_value = "AN000465",
  output_item = "SummarizedExperiment")

data_positive_mode <- do_query(
  context = "study",
  input_item = "analysis_id",
  input_value = "AN000464",
  output_item = "SummarizedExperiment")
```

The data have been saved into two R objects of class `SummarizedExperiment`. In order to have a more general dataset we continue the steps of the vignette.


# Standardize names

First the metabolite names are standardized- This is done using `rvest`package a specialized tool for web scrapping. There are easier and friendlier alternatives.

```{r}
metaboliteNamesURL <- "https://www.metabolomicsworkbench.org/data/show_metabolites_by_study.php?STUDY_ID=ST000291&SEARCH_TYPE=KNOWN&STUDY_TYPE=MS&RESULT_TYPE=1"
metaboliteNames <- metaboliteNamesURL %>% 
  read_html() %>% 
  html_nodes(".datatable")

metaboliteNames_negative <- metaboliteNames %>%
  .[[1]] %>%
  html_table() %>%
  dplyr::select(`Metabolite Name`, PubChemCompound_ID, `Kegg Id`)

metaboliteNames_positive <- metaboliteNames %>%
  .[[2]] %>%
  html_table() %>%
  dplyr::select(`Metabolite Name`, PubChemCompound_ID, `Kegg Id`)

metaboliteNames <- bind_rows(metaboliteNames_negative, metaboliteNames_positive) %>%
  dplyr::rename(names = 1, PubChem = 2, KEGG = 3) %>%
  mutate(KEGG = ifelse(KEGG == "-", "UNKNOWN", KEGG),
         PubChem = ifelse(PubChem == "-", "UNKNOWN", PubChem)) %>%
  filter(!duplicated(PubChem))
```

The names obtained are now assigned to the metabolites,

```{r}
require(SummarizedExperiment)
features_negative <- assay(data_negative_mode) %>%
  dplyr::slice(-n())
rownames(features_negative) <- rowData(data_negative_mode)$metabolite[1:(length(rowData(data_negative_mode)$metabolite)-1)]

## positive mode features
features_positive <- assay(data_positive_mode) %>%
  dplyr::slice(-n())
rownames(features_positive) <- rowData(data_positive_mode)$metabolite[1:(length(rowData(data_positive_mode)$metabolite)-1)]
```

Positive and negative values are combined into a unique `features`table.

Object about groups (metadata information) is stored into a single table.

Metabolite names are saved in a third table.

```{r}
features <- bind_rows(features_negative, features_positive) %>%
  tibble::rownames_to_column("names") %>%
  right_join(metaboliteNames, by = "names") %>%
  select(-names, -KEGG) %>%
  tibble::column_to_rownames("PubChem")

## metadata
pdata <- colData(data_negative_mode) %>% # or "data_positive_mode". They are equal
  as.data.frame() %>%
  tibble::rownames_to_column("ID") %>%
  mutate(Treatment = case_when(Treatment == "Baseline urine" ~ "Baseline",
                               Treatment == "Urine after drinking apple juice" ~ "Apple",
                               Treatment == "Urine after drinking cranberry juice" ~ "Cranberry"))
```

These three objects contain all the information that is needed foir further analyses.

```{r}
dim(assay(features))
assay(features)[1:10, c(1:5,30:35)]
```


```{r}
dim(pdata)
head(pdata)
```


```{r}
dim(metaboliteNames)
metaboliteNames[1:10,]
```

All three objkects are save into separate csv files using ";" as separator, to avoid possible conflicts with commas in metabolite names.

```{r}
write.table(assay(features), sep=";", file="features.csv")
write.table(pdata, sep=";", file="metadata.csv")
write.table(metaboliteNames, sep=";",
            file="metaboliteNames.csv")
```

In order to facilitate exploration to non bioinformatic users a single spreadsheet has been created with each object in one tab. Use it at your own risk.

```{r}
write.xlsx(assay(features), file="ST000291.xlsx", sheetName = "features", col.names = TRUE, row.names = FALSE, append = FALSE)
write.xlsx(pdata, file="ST000291.xlsx", sheetName = "metadata", col.names = TRUE, row.names = FALSE, append = TRUE)
write.xlsx(metaboliteNames, file="ST000291.xlsx", sheetName = "metaboliteNames", col.names = TRUE, row.names = TRUE, append = TRUE)
```

